Using Cell Phone Keyboards is (\NP) Hard

### 2010-01-15 article.tex
Uses LLNCS format, and the makefile has a clean target.

\documentclass{article}

\usepackage{amsmath,amsthm}
\usepackage{url}
\usepackage{graphicx}

\newcommand{\newword}[1]{{\bf #1}}
\newtheorem{problem}{Problem}
\newtheorem{theorem}{Theorem}
\newcommand{\NP}{\ensuremath{\mathcal{NP}}}

\title{Using Cell Phone Keyboards is (\NP) Hard}
\author{P. Boothe}

@@

Before we get any farther, let us sketch the basic problem that we will keep
revising and revisiting throughout this paper.
\begin{problem}[{\sc MinimumKeystrokes}]~\\
{\sc Instance}: A set of letters corresponding to an alphabet $A$ $(|A| =
n)$, a number of keys $k$, an input method $IN$, and a set of
tuples of words and frequencies $W$.  The frequencies in $W$ are integers,
and the words are made up of solely of elements of $A$.  We will
treat $IN$ as a function which, given a partition of $A$ and a word
$w$, returns how many keystrokes are required to type $w$.~\\
        ~\\
{\sc Question}: What is the best partition of $A$ into $k$ sets, such that the
total number of keystrokes to type every word in $W$ its associated frequency
times is minimized?  Equivalently, what is the partition of $P$ of $A$ ($|P| = k$) that
will minimize
$$\sum_{(w,f)\in W} f*IN(w,P)$$
\label{probtemplate}
\end{problem}

@@

\begin{problem}[{\sc BasicCellPhoneTyping}]~\\
{\sc Instance}: A set of letters corresponding to an alphabet $A$ $(|A| =
n)$, a number of keys $k$, and a set of
tuples of words and frequencies $W$.  The frequencies in $W$ are integers,
and the words are made up of solely of elements of $A$. \\
        ~\\
{\sc Question}: What is the best partition $P$ of $A$ into $k$ sequences, such
that the total number of button presses to type every word in $W$ its
associated frequency times is minimized?  The number of button presses is equal
to the order of the letter in the sequence assigned to a given key.  For
example, in the key $2\to\{abc\}$, $a$ requires one keystroke and $c$ requires
3.  Equivalently, if we denote the number of button presses required to type
letter $a$ with partition $P$ as $IN_P(a)$, then we want to find the $P$ that
minimizes $$\sum_{(w,f) \in W}\sum_{c\in w} IN_P(c) * f$$
\label{bcpt}
\end{problem}

Unlike almost every other problem dealing with key assignment, we can construct
an efficient algorithm for this problem.  We can construct a provable optimal
greedy algorithm which works in time $O(|W| + |A| \log |A|)$.  Our algorithm is
detailed in Figure~\ref{greedyalg}, and involves creating a histogram of the
letters, and then assigning letters to keys round-robin style in the order from
most-popular to least-popular.  To prove optimality, we invoke an exchange argument.

\begin{theorem}
The greedy algorithm finds an optimal solution to {\sc BasicCellPhoneTyping}.
\label{basicthm}
\end{theorem}
\begin{proof}
We will assume, for simplicity of proof, that all letters occur a different number of times in the corpus.  
We begin by comparing two assignments of letters to keys by, for each assignment,
constructing a sequence of sets, or spectrum, $S$.  The set $S_1$ (layer 1) is the set of all
letters which require a single button press (that is, they are the first
letter in their sequence on their assigned key), $S_2$ (layer 2) is the set of
all letters which require two button presses, and so on.  If two assignments
have the same sequence of sets, then one assignment may be transformed into
the other assignment simply by swapping letter between keys without
increasing or decreasing the total number of button pushes required to enter
the corpus of words.  If two sequences have equal spectra, then we say their corresponding assignments are isomorphic.

@@

In an effort to minimize the pain of cellphone keyboard typing, cellular
telephone manufacturers have created the ``T9'' input method, which attempts to
guess which letter (of the 3 or 4 possible) is intended when a user hits a
single key.  Enhancements of this input method also provide ``speculative
lookahead'' to report, at any given time, what word it is that the user is most
likely trying to enter and provide a completion.  Unfortunately, because there
are fewer keys than there are letters in the English alphabet, there are words
which have different spellings but the same input sequence.  As an example, in
the traditional mapping of keys to characters ($abc\to 2$, $def\to3$, etc),
both ``me'' and ``of'' have the input sequence ''63''.  Two words with the same
input sequence force the user to press a third button to cycle through the
possibilities in order from most likely (``of'') to least likely (``me'').
Even worse: ``home'', ``good'', ``gone'', ``hood'', ``hoof'', ``hone'', and
``goof'' all have the input sequence ``4663''.  If the * key is used as the
``next match'' button, then the user will actually have to type in
``4663******'' to type in the word ``goof''.

When two words have the same input sequence (neglecting the *'s at the end) then these words are \newword{t9onyms}.  When two or more words are t9onyms, then the less-popular words require extra keypresses, raising the expected number of keypresses to type in a random selection of words from the english dictionary.  In order to type a given word, one must press one button for every character in the word, followed by pressing the * key as many times as there are t9onyms which are more likely than the desired wored.  Because typing on a cellphone keyboard is already an unpleasant experience, we would like to minimize the expected number of keystrokes.  The number of keystrokes a word requires is equal to the number of letters in the word, plus the number of t9onyms which are more frequent than the word.  Formally, we define the {\sc MinimalKeystrokes} problem as:

\begin{theorem}[{\sc MinimumKeystrokes}]~\\
\label{thm:minstrokes}
{\sc Instance} An alphabet $A$, a set of words and their associated probabilities $W$, and a set of keys $K$ ($|A| > |K|$).\\
~\\
{\sc Question} What is the mapping of $A \to K$ which minimizes the expected number of characters to type a word from $W$?
\end{theorem}

@@

\begin{theorem}~\\
{\sc Instance} An alphabet $A$, a set of words and their associated probabilities $W$, a set of keys $K$ ($|A| > |K|$), and a number $f$.\\
~\\
{\sc Question} Is there a mapping of $A \to K$ in which the expected number of characters to type a word from $W$ is less than $f$?
\end{theorem}

@@

However, our arbitrary remappings of the keyboard may be quickly deemed unusable.  Therefore, we define a refinement of the problem, in which the characters are required to be kept in alphabetical order as a sequence, and our only choices are about how to divide the subsequence into key assignments.  The default layout can be described as the partition $abc|def|ghi|jkl|mno|pqrs|tuv|wxyz$.  Is this partition optimal?  In order to decide this, we define the problem {\sc MinimumKeystrokePartition} as:
\begin{theorem}[{\sc MinimumKeystrokePartition}]~\\
\label{thm:minpartition}
{\sc Instance} A sequence of letters $A = \{a_1, a_2, \ldots \}$, a set of words and their associated probabilities $W$, and a set of keys $K = \{2, 3, 4, \ldots \}$ ($|A| > |K|$).\\
~\\
{\sc Question} What is the mapping $f$ of $A \to K$ which minimizes the expected number of characters to type a word from $W$, and where $f(a_1) = 2$ and if $f(a_i) = k$, then either $f(a_{i+1}) = k$ or $f(a_{i+1}) = k+1$?
\end{theorem}

As before, the corresponding decision problem is
\begin{theorem}~\\
{\sc Instance} A sequence of letters $A = \{a_1, a_2, \ldots \}$, a set of words and their associated probabilities $W$, a set of keys $K = \{2, 3, 4, \ldots \}$ ($|A| > |K|$), and a parameter $e$.\\
~\\
{\sc Question} Is there a mapping $f$ of $A \to K$ in which the expected number of characters to type a word from $W$ is less than $e$, and where $f(a_1) = 2$ and if $f(a_i) = k$, then $f(a_{i+1}) \in \{k, k+1\}$?
\end{theorem}

>>>>

\documentclass[runningheads]{llncs}

\usepackage{amsmath}
\usepackage{url}
\usepackage{graphicx}
\usepackage{salgorithm}

\newcommand{\newword}[1]{{\bf #1}}
\newcommand{\NP}{\ensuremath{\mathcal{NP}}}
\spnewtheorem{prob}[theorem]{Problem}{\bfseries}{\itshape}

\title{Using Cell Phone Keyboards is (\NP) Hard}
\author{Peter Boothe}
\institute{Manhattan College\\
\email{peter.boothe@manhattan.edu}
}

@@

Before we get any farther, let us sketch the basic problem that we will keep
revising and revisiting throughout this paper.
\begin{prob}[{\sc MinimumKeystrokes}]~\\
{\sc Instance}: A set of letters corresponding to an alphabet $A$ $(|A| =
n)$, a number of keys $k$, an input method $IN$, and a set of
tuples of words and frequencies $W$.  The frequencies in $W$ are integers,
and the words are made up of solely of elements of $A$.  We will
treat $IN$ as a function which, given a partition of $A$ and a word
$w$, returns how many keystrokes are required to type $w$.~\\
        ~\\
{\sc Question}: What is the best partition of $A$ into $k$ sets, such that the
total number of keystrokes to type every word in $W$ its associated frequency
times is minimized?  Equivalently, what is the partition of $P$ of $A$ ($|P| = k$) that
will minimize
$$\sum_{(w,f)\in W} f*IN(w,P)$$
\label{probtemplate}
\end{prob}

@@

\begin{prob}[{\sc BasicCellPhoneTyping}]~\\
{\sc Instance}: A set of letters corresponding to an alphabet $A$ $(|A| =
n)$, a number of keys $k$, and a set of
tuples of words and frequencies $W$.  The frequencies in $W$ are integers,
and the words are made up of solely of elements of $A$. \\
        ~\\
{\sc Question}: What is the best partition $P$ of $A$ into $k$ sequences, such
that the total number of button presses to type every word in $W$ its
associated frequency times is minimized?  The number of button presses is equal
to the order of the letter in the sequence assigned to a given key.  For
example, in the key $2\to\{abc\}$, $a$ requires one keystroke and $c$ requires
3.  Equivalently, if we denote the number of button presses required to type
letter $a$ with partition $P$ as $IN_P(a)$, then we want to find the $P$ that
minimizes $$\sum_{(w,f) \in W}\sum_{c\in w} IN_P(c) * f$$
\label{bcpt}
\end{prob}

\begin{figure}
\begin{algorithm}
\Algorithm{GreedyBasicCellPhoneTyping}~$(A, k, W)$\+\\
    // $A$ is the set of letters \\
    // $k$ is the number of keys \\
    // $W$ is a set of pairs of words and their corresponding frequencies \\
    lettercount $\gets$ new\_map() \\
    \For $c \in A$\+\\
        lettercount$[c] \gets 0$\-\\
    \For (word, frequency) $\in W$\+\\
        \For $c \in$ word \+\\
            lettercount$[c] \gets $ lettercount$[c] +$frequency\-\-\\
    ordered $\gets \left[ ({\rm lettercount}[c], c)~\For c \in {\rm lettercount} \right]$\\
    sort(ordered)\\
    keys $\gets$ a array of length $k$ where each element is an empty list\\
    \For $i \gets 0 \ldots length($ordered$)$\+\\
        (count, char) $\gets$ ordered$[i]$\\
        append(char, keys$[i mod k]$)\-\\
    \Return keys
\end{algorithm}
\caption{The greedy algorithm for Problem~\ref{bcpt}.}
\label{greedyalg}
\end{figure}

To solve {\sc GreedyBasicCellPhoneTyping} we construct a provably optimal
greedy algorithm which works in time $O(|W| + |A| \log |A|)$.  Our algorithm is
detailed in Figure~\ref{greedyalg}, and involves creating a histogram of the
letters, and then assigning letters to keys round-robin style in the order from
most-popular to least-popular.  To prove optimality, we invoke an exchange argument.

\begin{theorem}
The greedy algorithm finds an optimal solution to {\sc BasicCellPhoneTyping}.
\label{basicthm}
\end{theorem}
\begin{proof}
We will assume, for simplicity of proof, that all letters occur a different number of times in the corpus.  
We begin by comparing two assignments of letters to keys by, for each assignment,
constructing a sequence of sets, or spectrum, $S$.  The set $S_1$ (layer 1) is the set of all
letters which require a single button press (that is, they are the first
letter in their sequence on their assigned key), $S_2$ (layer 2) is the set of
all letters which require two button presses, and so on.  If two assignments
have equal spectra, then one assignment may be transformed into
the other assignment simply by swapping letter between keys without
increasing or decreasing the total number of button pushes required to enter
the corpus of words.   Two assignments are isomorphic iff their spectra are equal.

@@

In an effort to minimize the pain of cellphone keyboard typing, cellular
telephone manufacturers have created the T9 input method, which attempts to
guess which letter (of the 3 or 4 possible) is intended when a user hits a
single key.  Enhancements of this input method also provide \newword{speculative
lookahead} to report, at any given time, what word it is that the user is most
likely trying to enter and provide a completion.  Unfortunately, because there
are fewer keys on the cellphone keyboard than there are letters in the English alphabet, there are words
which have different spellings but the same input sequence.  As an example, in
the traditional mapping of keys to characters
both ``me'' and ``of'' have the input sequence ''63''.  Two words with the same
input sequence force the user to press a third button to cycle through the
possibilities in order from most likely (``of'') to least likely (``me'').
Even worse: ``home'', ``good'', ``gone'', ``hood'', ``hoof'', ``hone'', and
``goof'' all have the input sequence ``4663''.  If the * key is used as the
``next match'' button, then the user will actually have to type in
``4663******'' to type in the word ``goof'', for a total of 10 button presses --- exactly the same number of button presses required using the default layout and the basic input method, and three more button presses than is required when using the basic input method with an optimized layout.

When two words have the same input sequence (neglecting the *'s at the end) then these words are \newword{t9onyms}.  When two or more words are t9onyms, then the less-popular words require extra keypresses, raising the expected number of keypresses to type in our corpus.  In order to type a given word, one must press one button for every character in the word, followed by pressing the * key as many times as there are t9onyms which are more likely than the desired wored.  Because typing on a cellphone keyboard is already an unpleasant experience, we would like to minimize the expected number of keystrokes.  The number of keystrokes a word requires is equal to the number of letters in the word, plus the number of t9onyms which are more frequent than the word.  Formally, we extend Problem~\ref{probtemplate} and define the {\sc MinimumT9Keystrokes} problem as:

\begin{prob}[{\sc MinimumT9Keystrokes}]~\\
\label{thm:minstrokes}
{\sc Instance} An alphabet $A$, a set of words and their associated frequencies $W$, and a number of keys $k$ ($|A| > k$).\\
~\\
{\sc Question} What is the partion $P$ of $A$ into $k$ sets which minimizes the total button presses required to enter the entire corpus using the T9 input method?
\end{prob}

@@

\begin{prob}~\\
{\sc Instance} An alphabet $A$, a set of words and their associated frequencies $W$, a number of keys $k$ ($|A| > k$), and a number $t$.\\
~\\
{\sc Question} Is there a partition of $A$ into $k$ sets in which the total number of keystrokes required to enter the entire corpus using the T9 input method requires no more than $t$ button presses?
\end{prob}

@@

However, our arbitrary remappings of the keyboard may be quickly deemed unusable.  Therefore, we define a refinement of the problem, in which the characters are required to be kept in alphabetical order as a sequence, and our only choices are about how to divide the subsequence into key assignments.  The default layout can be described as the partition $abc|def|ghi|jkl|mno|pqrs|tuv|wxyz$.  Is this partition optimal?  In order to decide this, we define the problem {\sc MinimumKeystrokePartition} as:
\begin{prob}[{\sc MinimumKeystrokePartition}]~\\
\label{thm:minpartition}
{\sc Instance} A sequence of letters $A = \{a_1, a_2, \ldots \}$, a set of words and their associated probabilities $W$, and a set of keys $K = \{2, 3, 4, \ldots \}$ ($|A| > |K|$).\\
~\\
{\sc Question} What is the mapping $f$ of $A \to K$ which minimizes the expected number of characters to type a word from $W$, and where $f(a_1) = 2$ and if $f(a_i) = k$, then either $f(a_{i+1}) = k$ or $f(a_{i+1}) = k+1$?
\end{prob}

As before, the corresponding decision problem is
\begin{prob}~\\
{\sc Instance} A sequence of letters $A = \{a_1, a_2, \ldots \}$, a set of words and their associated probabilities $W$, a set of keys $K = \{2, 3, 4, \ldots \}$ ($|A| > |K|$), and a parameter $e$.\\
~\\
{\sc Question} Is there a mapping $f$ of $A \to K$ in which the expected number of characters to type a word from $W$ is less than $e$, and where $f(a_1) = 2$ and if $f(a_i) = k$, then $f(a_{i+1}) \in \{k, k+1\}$?
\end{prob}
